---
title: "Tangi across three UTMBs"
subtitle: "Where a 2026-calibre run by Tangi Taton would have landed in the 2024 and 2025 editions — and how much of the answer is really knowable."
author: "Alexis Konarski"
date: today
format:
  html:
    theme: ft.scss
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: false
    code-fold: false
    fig-width: 9
    fig-height: 6
    fig-dpi: 160
    embed-resources: true
    df-print: kable
execute:
  warning: false
  message: false
  echo: false
---

```{r setup}
#| include: false
library(readr)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(scales)
library(ggrepel)
library(stringr)
library(knitr)
library(ggtext)
library(splines)

source("ft_style.R")
knitr::opts_chunk$set(dev = "png", dpi = 160, echo = FALSE)
theme_set(theme_ft())

set.seed(20260907)

spotlight_name <- "Tangi TATON"
spotlight_bib  <- 535L

ord <- function(n) paste0(n, dplyr::case_when(
  n %% 100 %in% 11:13 ~ "th", n %% 10 == 1 ~ "st",
  n %% 10 == 2 ~ "nd", n %% 10 == 3 ~ "rd", .default = "th"))
fmt_hms <- function(s) {
  h <- floor(s / 3600); m <- floor((s %% 3600) / 60)
  sprintf("%d:%02d", h, m)
}
mins <- function(h) paste0(round(abs(h) * 60), " min")
# Every chart in this report reads all three editions.
srcline <- function(note = NULL) ft_source(note, editions = "2024–2026 editions")
key <- function(x) toupper(trimws(x))     # join key for runners across editions
```

```{r load}
# Full field (finishers + DNFs) for each edition. 2026 is the existing file;
# 2024 / 2025 are scraped by R/scrape_history.R.
paths <- c(
  `2024` = "../data/processed/runners_all_utmb_2024.csv",
  `2025` = "../data/processed/runners_all_utmb_2025.csv",
  `2026` = "../data/processed/runners_all_utmb.csv"
)

# The index range over which the index->time line is well measured in every
# edition: high enough that few runners are lost to the cut-off, low enough
# that the sample is not just the elite. See the methodology.
FIT_LO <- 650; FIT_HI <- 900
in_range <- function(d) filter(d, utmb_index >= FIT_LO, utmb_index <= FIT_HI)

prep <- function(df) {
  s <- df |>
    filter(outcome %in% c("Finisher", "DNF"), !is.na(utmb_index)) |>
    mutate(time_h = race_time_seconds / 3600,
           fin    = as.integer(outcome == "Finisher"))
  f <- s |>
    filter(outcome == "Finisher", !is.na(race_time_seconds)) |>
    arrange(race_time_seconds) |>
    mutate(fin_rank = row_number())
  list(
    s = s, f = f,
    n_start = nrow(s), n_fin = nrow(f), n_dnf = sum(s$outcome == "DNF"),
    dnf_rate = sum(s$outcome == "DNF") / nrow(s),
    med_index = median(s$utmb_index),
    fit      = lm(time_h ~ utmb_index, data = in_range(f)),   # primary
    fit_full = lm(time_h ~ utmb_index, data = f)              # sensitivity
  )
}

ed <- map(paths, \(p) prep(read_csv(p, show_col_types = FALSE)))
y24 <- ed[["2024"]]; y25 <- ed[["2025"]]; y26 <- ed[["2026"]]

# Seeds. Two bases, and the distinction matters more than anything else here:
#   seed_start = rank among everyone who STARTED rated above you
#   seed_fin   = rank among everyone who FINISHED rated above you
seed_start <- function(Y, I) sum(Y$s$utmb_index > I) + 1L
seed_fin   <- function(Y, I) sum(Y$f$utmb_index > I) + 1L
par_time   <- function(Y, I, full = FALSE)
  as.numeric(predict(if (full) Y$fit_full else Y$fit, data.frame(utmb_index = I)))
```

```{r baseline}
TI           <- 758L                                   # his 2026 UTMB Index
tangi26      <- y26$f |> filter(bib == spotlight_bib)
tangi_time_s <- tangi26$race_time_seconds
tangi_time_h <- tangi26$time_h
tangi_rank26 <- tangi26$fin_rank
tangi_cat    <- tangi26$category

seedS26 <- seed_start(y26, TI)   # 229th on the start list
seedF26 <- seed_fin(y26, TI)     # 133rd among those who finished

# Form, measured LIKE FOR LIKE: seed and rank both on the finisher basis.
formF_delta <- seedF26 - tangi_rank26      # -20
formF_ratio <- tangi_rank26 / seedF26      # 1.15
# The mixed-basis version the first draft of this report used, kept only to
# show why it is wrong.
formS_delta <- seedS26 - tangi_rank26      # +76, an artefact

# Personal margin against the index->time line (primary = restricted fit).
pred26  <- par_time(y26, TI)
resid26 <- tangi_time_h - pred26
pred26_full  <- par_time(y26, TI, full = TRUE)
resid26_full <- tangi_time_h - pred26_full
resid_sd26   <- sd(resid(y26$fit))
```

```{r form-population}
# Both versions of "form" for every 2026 finisher, so Tangi can be placed in
# the distribution rather than judged against zero.
popform <- y26$f |>
  mutate(seedS = map_int(utmb_index, \(i) seed_start(y26, i)),
         seedF = map_int(utmb_index, \(i) seed_fin(y26, i)),
         dS = seedS - fin_rank, dF = seedF - fin_rank,
         rhoS = fin_rank / seedS, rhoF = fin_rank / seedF)
tp <- popform |> filter(bib == spotlight_bib)

pct_dS   <- mean(popform$dS < tp$dS) * 100          # his percentile on the artefact
pct_dF   <- mean(popform$dF < tp$dF) * 100          # his percentile like-for-like
med_dS   <- median(popform$dS); med_dF <- median(popform$dF)
attrition_gain <- tp$dS - tp$dF                     # places purely from DNFs above him

# A2: rank is not bounded by higher-rated finishers.
ahead26      <- y26$f |> filter(fin_rank < tangi_rank26)
ahead_weaker <- sum(ahead26$utmb_index <= TI)
```

```{r drift}
# Index scale drift, measured within runner on those who ran two editions.
returnees <- function(yr) {
  a <- ed[[yr]]$s |> mutate(k = key(fullname)) |> distinct(k, .keep_all = TRUE) |>
    select(k, i_old = utmb_index)
  b <- y26$s |> mutate(k = key(fullname)) |> distinct(k, .keep_all = TRUE) |>
    select(k, i_new = utmb_index)
  inner_join(a, b, by = "k") |> mutate(d = i_new - i_old)
}
ret <- map(c(`2024` = "2024", `2025` = "2025"), returnees)
drift <- map_dbl(ret, \(m) median(m$d))
n_ret <- map_int(ret, nrow)
pct_identical <- map_dbl(ret, \(m) 100 * mean(m$i_old == m$i_new))

# His own contemporaneous rating on UTMB day 2025, supplied outside the API.
TI_2025_actual <- 732L
```

```{r uniform-test}
# Is the year effect a pure level shift? If so the index->time slope should not
# move between editions. Tested on the fitted range and on the whole field.
int_p <- function(restrict) {
  d <- imap(ed, \(Y, yr) (if (restrict) in_range(Y$f) else Y$f) |> mutate(year = yr)) |>
    list_rbind()
  anova(lm(time_h ~ utmb_index + year, data = d),
        lm(time_h ~ utmb_index * year, data = d))[["Pr(>F)"]][2]
}
int_p_restricted <- int_p(TRUE)
int_p_full       <- int_p(FALSE)
```

```{r projections}
# Four lenses. A and B are naive bookends; C and D are the working ones.
project <- function(Y, yr) {
  sS <- seed_start(Y, TI); sF <- seed_fin(Y, TI)
  # Drift-adjusted seed: his 2026-scale index expressed on that year's scale.
  I_adj  <- TI - (if (yr == "2026") 0 else drift[[yr]])
  sF_adj <- seed_fin(Y, I_adj)
  predY      <- par_time(Y, TI)
  predY_full <- par_time(Y, TI, full = TRUE)
  tibble(
    year = yr,
    n_start = Y$n_start, n_fin = Y$n_fin, dnf_rate = Y$dnf_rate,
    med_index = Y$med_index,
    seed_start = sS, seed_fin = sF, seed_fin_adj = sF_adj,
    raw_time = sum(Y$f$race_time_seconds < tangi_time_s) + 1L,      # B
    adj_delta = pmax(1L, sF - formF_delta),                          # C
    adj_ratio = pmax(1L, as.integer(round(sF * formF_ratio))),       # C
    adj_drift = pmax(1L, sF_adj - formF_delta),                      # C, drift-adjusted
    model      = sum(Y$f$time_h < (predY + resid26)) + 1L,           # D
    model_full = sum(Y$f$time_h < (predY_full + resid26_full)) + 1L, # D, full-range fit
    par_h = predY, par_h_full = predY_full, proj_time_h = predY + resid26
  )
}
proj <- imap(ed, \(Y, yr) project(Y, yr)) |> list_rbind()

# The central estimate is the mean of the three working projections
# (C-delta, C-ratio, D), and the reported range is their spread.
proj <- proj |>
  rowwise() |>
  mutate(central   = round(mean(c_across(c(adj_delta, adj_ratio, model)))),
         spread_lo = min(c_across(c(adj_delta, adj_ratio, model))),
         spread_hi = max(c_across(c(adj_delta, adj_ratio, model)))) |>
  ungroup()

g <- \(yr, col) proj[[col]][proj$year == yr]
```

```{r uncertainty}
prior <- proj |> filter(year != "2026")
b <- \(yr, col) prior[[col]][prior$year == yr]

# What if he had run exactly to his rating instead of his 2026 margin? The
# cleanest sensitivity to the one thing being transplanted.
par_rank <- map_int(c(`2024` = "2024", `2025` = "2025"),
                    \(yr) sum(ed[[yr]]$f$time_h < g(yr, "par_h")) + 1L)

# The widest defensible region per year: the sampling range of the fits, the
# spread across methods, and the zero-margin scenario, rounded outward to 5.
# Sampling error in the two fitted lines behind lens D.
boot_D <- map(c("2024", "2025"), \(yr) {
  v <- map_int(1:600, \(i) {
    a <- in_range(slice_sample(y26$f, prop = 1, replace = TRUE))
    c2 <- in_range(slice_sample(ed[[yr]]$f, prop = 1, replace = TRUE))
    m26 <- lm(time_h ~ utmb_index, a); my <- lm(time_h ~ utmb_index, c2)
    e <- tangi_time_h - predict(m26, data.frame(utmb_index = TI))
    sum(ed[[yr]]$f$time_h < (predict(my, data.frame(utmb_index = TI)) + e)) + 1L
  })
  tibble(year = yr, lo = round(quantile(v, .025)), hi = round(quantile(v, .975)))
}) |> list_rbind()
bd <- \(yr, col) boot_D[[col]][boot_D$year == yr]

# Rank density around the projection: how many places a 15-minute error is worth.
dens15 <- map_int(c(`2024` = "2024", `2025` = "2025"),
                  \(yr) sum(abs(ed[[yr]]$f$time_h - g(yr, "proj_time_h")) < 0.25))

# The widest defensible region per year: the sampling range of the fits, the
# spread across methods and the zero-margin scenario, rounded outward to 5.
firm <- function(yr) {
  v <- c(bd(yr, "lo"), bd(yr, "hi"), b(yr, "spread_lo"), b(yr, "spread_hi"),
         par_rank[[yr]])
  c(lo = floor(min(v) / 5) * 5, hi = ceiling(max(v) / 5) * 5)
}
gain <- tangi_rank26 - prior$central
```

# The question

Tangi Taton (bib `r spotlight_bib`, UTMB Index **`r TI`**) finished **UTMB 2026** in
**`r fmt_hms(tangi_time_s)`**, **`r ord(tangi_rank26)`** of `r comma(y26$n_fin)` finishers
(category `r tangi_cat`). How would the *same runner, running equally well*, have placed in the
**2024** and **2025** editions?

Neither finish time nor raw rank answers this: each year a different field turns up on a different
day. The decisive fact is that **2026 was an unusually deep edition at his level** —
`r sum(y26$f$utmb_index > TI)` higher-rated runners reached the finish, against
`r sum(y24$f$utmb_index > TI)` in 2024 and `r sum(y25$f$utmb_index > TI)` in 2025. The same run is
worth fewer places when more strong runners are in the way, so ranks have to be adjusted for field
strength before they can be compared across years.

The question is a conditional one: *if he had run the same race relative to his rating*, what would
that have been worth in a thinner field? "Running equally well" is an assumption imposed here, not a
prediction — nobody knows what he would actually have done on a different mountain day. What follows
works that assumption through, and then sets out how much weight it can bear.

# How would he have finished? {#headline}

```{r headline-table}
prior |>
  transmute(
    Edition = year,
    Finishers = comma(n_fin),
    `Higher-rated finishers` = seed_fin - 1L,
    `Projected place` = ord(central),
    `Across the three methods` = paste0(ord(spread_lo), "–", ord(spread_hi)),
    `Places gained on his 153rd` = paste0("+", tangi_rank26 - central)
  ) |>
  kable(align = "lrrrrr")
```

The same run puts him around **`r ord(g("2024","central"))` in 2024** and
**`r ord(g("2025","central"))` in 2025** — roughly `r tangi_rank26 - g("2024","central")` and
`r tangi_rank26 - g("2025","central")` places ahead of his real `r ord(tangi_rank26)`. Three
different ways of doing the arithmetic ([set out below](#the-lenses)) land within
`r max(prior$spread_hi - prior$spread_lo)` places of each other.

That closeness is reassuring about the arithmetic and nothing more. The three methods share one
input — his single 2026 race — so they cannot disagree about it, and **their spread is not an
uncertainty interval**. What can be said about precision is [set out further down](#firm); the short
version is that the place should be read to the nearest ten or twenty, not to the unit.
`r dens15[["2024"]]` finishers sit within a quarter of an hour of his projected 2024 time and
`r dens15[["2025"]]` within a quarter of an hour in 2025, so a 15-minute error in the modelling is
worth a dozen places on its own.

```{r headline-chart}
#| fig-height: 4.6
#| fig-cap: "The projected place in each prior edition, against Tangi's real 2026 finish. Further right is better."
hc <- prior |>
  left_join(boot_D, by = "year") |>
  mutate(year = factor(year, levels = c("2025", "2024")))

ggplot(hc, aes(y = year)) +
  geom_linerange(aes(xmin = lo, xmax = hi), linewidth = 7, colour = ft_paper2) +
  geom_linerange(aes(xmin = spread_lo, xmax = spread_hi), linewidth = 2.4, colour = ft_neutral) +
  geom_vline(xintercept = tangi_rank26, colour = ft_ink2, linetype = "22", linewidth = 0.4) +
  geom_point(aes(x = central), size = 4, colour = ft_blue) +
  geom_text(aes(x = central, label = ord(central)), vjust = -1.4,
            family = "Helvetica Neue", colour = ft_ink, size = 3.4) +
  annotate("text", x = tangi_rank26, y = 2.42, label = "his real 2026 finish  ",
           hjust = 1, family = "Helvetica Neue", colour = ft_ink2, size = 3.1) +
  scale_x_reverse(limits = c(max(hc$hi) + 25, 0)) +
  labs(title = paste0("The same run is worth ", min(gain), "–", max(gain),
                      " more places in the thinner fields"),
       subtitle = paste0("Projected place (dot), the spread across the three methods (dark bar) ",
                         "and the sampling range of the fitted lines (pale bar)."),
       x = "Projected finishing place (reversed — further right is better)", y = NULL,
       caption = srcline("Neither bar is a full uncertainty interval — see 'How firm is this?'"))
```

# What his 2026 run was actually worth {#form}

Before his run can be carried anywhere it has to be valued correctly — and the obvious way of valuing
it is wrong.

The obvious way is to compare where he was seeded with where he finished. On the 2026 start list a
`r TI` index put him **`r ord(seedS26)`**; he finished **`r ord(tangi_rank26)`**, apparently
*beating* his seed by `r formS_delta` places. That comparison is broken, and the way it is broken
flatters every finisher.

The seed counts everyone rated above him who **started**; the rank counts only those who
**finished**. `r comma(y26$n_dnf)` runners left the denominator without leaving the numerator, so
every finisher shows a positive margin by construction. Across all `r comma(nrow(popform))` 2026
finishers the median "margin" is **`r med_dS` places**; Tangi's `r formS_delta` is only the
**`r ord(round(pct_dS))` percentile** of it.

Put both sides on the same basis — seed among finishers, rank among finishers — and the margin
reverses:

```{r form-table}
tibble(
  Basis = c("Seed on the start list (mixed basis)", "Seed among finishers (like for like)"),
  Seed = c(ord(seedS26), ord(seedF26)),
  Finish = ord(tangi_rank26),
  Margin = c(paste0("+", formS_delta, " places"), paste0(formF_delta, " places")),
  `Percentile of the field` = c(paste0(ord(round(pct_dS))), paste0(ord(round(pct_dF))))
) |> kable(align = "lrrrr")
```

Of that apparent +`r formS_delta`-place margin, **`r attrition_gain` places came from higher-rated runners
dropping out ahead of him** — attrition, not racing. On a like-for-like basis he finished
**`r abs(formF_delta)` places behind his rating** among those who finished, against a field median of
**+`r med_dF`**. He ran an unremarkable race for his rating: solidly, but not above it.

```{r form-chart}
#| fig-height: 4.4
#| fig-cap: "The distribution of the like-for-like margin (seed among finishers minus finishing rank) across all 2026 finishers. Positive = finished ahead of rating."
ggplot(popform, aes(dF)) +
  geom_histogram(binwidth = 40, fill = ft_neutral, colour = NA) +
  geom_vline(xintercept = med_dF, colour = ft_ink2, linetype = "22", linewidth = 0.4) +
  geom_vline(xintercept = tp$dF, colour = ft_blue, linewidth = 0.8) +
  annotate("text", x = tp$dF, y = Inf, label = paste0("  Tangi: ", tp$dF),
           hjust = 0, vjust = 1.6, family = "Helvetica Neue", colour = ft_blue, size = 3.4) +
  annotate("text", x = med_dF, y = Inf, label = paste0("field median: +", med_dF, "  "),
           hjust = 1, vjust = 1.6, family = "Helvetica Neue", colour = ft_ink2, size = 3.4) +
  labs(title = "On a like-for-like basis his 2026 run was an average one for his rating",
       subtitle = paste0("Places gained on the finisher-basis seed, all ", comma(nrow(popform)),
                         " finishers carrying an index. He is at the ", ord(round(pct_dF)),
                         " percentile."),
       x = "Places gained on seed (finisher basis)", y = "Finishers",
       caption = srcline())
```

This does not change the direction of the answer, but it changes what is being carried across the
years. What gets transplanted below is **an average run for a `r TI` index**, not an outstanding one,
and the reason he places better in 2024 and 2025 is entirely that the field was thinner — not that he
had a special day.

# The lenses

Four lenses, from naïve to working. Every projected rank is a rank **among finishers**, and every
seed is counted **among finishers**, so the two are always on the same basis.

**A · Seed — "on paper".**
Count the higher-rated runners in each year's field. Among finishers that is
**`r ord(g("2024","seed_fin"))`** (2024), **`r ord(g("2025","seed_fin"))`** (2025) and
**`r ord(seedF26)`** (2026). This is expectation, not result, but it is the crux of the question:
same runner, better starting position, because fewer strong runners were in his way.

**B · Raw time — the naïve answer.**
Slot `r fmt_hms(tangi_time_s)` straight into each year's finish list:
`r ord(g("2024","raw_time"))` in 2024, `r ord(g("2025","raw_time"))` in 2025. Distrust it — it
assumes identical conditions across years, when a typical `r TI`-index finisher needed
`r mins(g("2024","par_h") - pred26)` longer in 2024 than in 2026. Its swing measures the size of the
year effect, not a projection.

**C · Seed-adjusted — his rating, plus his form.**
Carry his measured like-for-like form ($\delta_F = `r formF_delta`$ places, or equivalently
$\rho_F = `r round(formF_ratio, 2)`\times$ his seed) onto each year's own finisher-basis seed. That
gives **`r ord(g("2025","adj_delta"))`–`r ord(g("2025","adj_ratio"))` in 2025** and
**`r ord(g("2024","adj_delta"))`–`r ord(g("2024","adj_ratio"))` in 2024**.

**D · Index-residual time model.**
Fit each edition's UTMB Index → finish-time line over the well-measured index range
(`r FIT_LO`–`r FIT_HI`). In 2026 a typical `r TI`-index finisher took **`r fmt_hms(pred26 * 3600)`**;
Tangi ran **`r mins(resid26)` `r if (resid26 < 0) "faster" else "slower"`**. Holding that margin and
reading it against each year's own line puts him at **`r ord(g("2025","model"))` in 2025** and
**`r ord(g("2024","model"))` in 2024**.

C and D are **not independent**. Both are anchored on the same single run and the same index; they
differ only in whether the anchor is expressed in places or in minutes. Their agreement is
reassuring about arithmetic, not about the underlying assumption.

```{r lens-table}
proj |>
  transmute(
    Edition = year,
    `A · seed` = ord(seed_fin),
    `B · raw time` = ord(raw_time),
    `C · delta` = ord(adj_delta),
    `C · ratio` = ord(adj_ratio),
    `D · time model` = ord(model),
    `D · full-range fit` = ord(model_full)
  ) |>
  kable(align = "lrrrrrr",
        caption = "Every lens, every edition. The 2026 row is a self-check: C reproduces his actual 153rd by construction, and D lands on it because his residual is measured there.")
```

```{r model-chart}
#| fig-cap: "Each edition's UTMB Index → finish-time line, fit over the well-measured index range. The vertical line marks Tangi's index; the points are his projected times once his 2026 margin is carried across."
grid_idx <- tibble(utmb_index = seq(FIT_LO, FIT_HI, by = 5))
lines <- imap(ed, \(Y, yr) grid_idx |>
                mutate(year = yr,
                       time_h = as.numeric(predict(Y$fit, newdata = grid_idx)))) |>
  list_rbind()
pts <- proj |> transmute(year, utmb_index = TI, time_h = proj_time_h)

ggplot(lines, aes(utmb_index, time_h, colour = year)) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = TI, colour = ft_ink2, linetype = "22", linewidth = 0.4) +
  geom_point(data = pts, size = 3.2) +
  geom_text_repel(data = pts, aes(label = paste0(year, ": ", fmt_hms(time_h * 3600))),
                  family = "Helvetica Neue", size = 3.3, seg.colour = ft_neutral,
                  direction = "y", nudge_x = 25, show.legend = FALSE) +
  scale_colour_manual(values = c(`2024` = ft_claret, `2025` = ft_orange, `2026` = ft_blue),
                      name = "Edition") +
  scale_y_continuous(labels = \(h) paste0(h, "h")) +
  labs(title = "The same index buys a different finish time each year",
       subtitle = paste0("Index→time fit per edition, index ", FIT_LO, "–", FIT_HI,
                         ". Points carry his 2026 margin (", mins(resid26), " ",
                         if (resid26 < 0) "faster" else "slower", " than par) across."),
       x = "UTMB Index", y = "Finish time",
       caption = srcline("Higher line = a slower edition at a given index")) +
  theme(legend.position = "top")
```

# How firm is this? {#firm}

Four things bound how precisely this can be answered. None of them changes the direction; all of
them argue against quoting a place to the unit.

**The margin being carried is thin.** Everything rests on one number: that Tangi ran
`r mins(resid26)` faster than a typical `r TI`-index finisher in 2026. Against a residual spread of
`r mins(resid_sd26)` around that line, this is a **z of `r round(resid26 / resid_sd26, 2)`** — about
a fifth of a standard deviation. Measured on a single race, that is close to indistinguishable from
an ordinary day, so the honest reading is not "he has a `r mins(resid26)` edge" but "he ran roughly
to his rating, perhaps a shade better". Carry *exactly* his rating instead — set the margin to zero —
and he lands `r ord(par_rank[["2024"]])` in 2024 and `r ord(par_rank[["2025"]])` in 2025, some
`r par_rank[["2024"]] - g("2024","model")` and `r par_rank[["2025"]] - g("2025","model")` places
back. That is the size of the entire personal-margin term.

**The lines themselves are estimates.** Lens D reads his margin against two fitted index→time lines,
each carrying sampling error of its own. Bootstrapping both fits (600 replicates) puts him at
`r ord(bd("2024","lo"))`–`r ord(bd("2024","hi"))` in 2024 and
`r ord(bd("2025","lo"))`–`r ord(bd("2025","hi"))` in 2025 — already several times wider than the
three methods disagree, and covering only one source of error.

**Places are cheap where he finishes.** `r dens15[["2024"]]` finishers sat within a quarter of an
hour of his projected 2024 time and `r dens15[["2025"]]` within a quarter of an hour in 2025. Fifteen
minutes is well inside the noise of any of this, and it is worth a dozen places.

**And he might not have finished at all.** `r percent(g("2024","dnf_rate"), 1)` of 2024 starters and
`r percent(g("2025","dnf_rate"), 1)` of 2025 starters did not, both above 2026's
`r percent(g("2026","dnf_rate"), 1)`. Every number here is conditional on him reaching Chamonix.

Read together, the defensible answer is a region rather than a place: **roughly
`r ord(firm("2024")[["lo"]])`–`r ord(firm("2024")[["hi"]])` in 2024 and
`r ord(firm("2025")[["lo"]])`–`r ord(firm("2025")[["hi"]])` in 2025**, comfortably ahead of
`r ord(tangi_rank26)` in both — and the reason is the field rather than the runner.

# Why the years differ: field and attrition {#why}

Two forces decide whether the same run is worth more or fewer places: **how deep the field is at his
level**, and **how brutal the day is**. Both worked against him in 2026.

```{r depth-table}
depth <- imap(ed, \(Y, yr) tibble(
  Edition = yr,
  `Higher-rated starters` = sum(Y$s$utmb_index > TI),
  `Higher-rated finishers` = sum(Y$f$utmb_index > TI),
  `Starters ≥ 800` = sum(Y$s$utmb_index >= 800),
  `Starters ≥ 850` = sum(Y$s$utmb_index >= 850),
  `Starters ≥ 900` = sum(Y$s$utmb_index >= 900)
)) |> list_rbind()
kable(depth, align = "lrrrrr",
      caption = "Field depth at and above Tangi's index (758). The first two columns are the same population on the two bases used throughout; the last three are start-list counts.")
```

In 2026, **`r sum(y26$f$utmb_index > TI)`** higher-rated runners reached the finish; in 2024 and
2025 only **`r sum(y24$f$utmb_index > TI)`** and **`r sum(y25$f$utmb_index > TI)`** did. The elite
tail says the same — `r sum(y26$s$utmb_index >= 900)` starters rated ≥ 900 in 2026 against
`r sum(y24$s$utmb_index >= 900)` and `r sum(y25$s$utmb_index >= 900)`. Tangi did not get slower; the
company got faster.

That said, a thin field above him is not a floor under his rank. **In 2026,
`r ahead_weaker` of the `r nrow(ahead26)` finishers ahead of him (`r percent(ahead_weaker/nrow(ahead26), 1)`)
were rated at or below him.** Finishing rank is not bounded by higher-rated finishers: to be
`r ord(tangi_rank26)` in 2024 he would need `r tangi_rank26 - 1L - (g("2024","seed_fin") - 1L)`
lower-rated finishers ahead of him out of `r comma(y24$n_fin - (g("2024","seed_fin") - 1L))` available,
which is entirely possible. A thin field above him makes a better placing the *expectation*, not a
certainty — which is why the projected place is best read as a region.

```{r context-table}
proj |>
  transmute(
    Edition = year,
    Starters = comma(n_start),
    Finishers = comma(n_fin),
    `DNF rate` = percent(dnf_rate, 1),
    `Median index` = round(med_index),
    `Par time at 758` = fmt_hms(par_h * 3600)
  ) |>
  kable(align = "lrrrrr")
```

```{r context-chart}
#| fig-height: 4.2
#| fig-cap: "Left: share of starters who did not finish. Right: the finish time a typical 758-index runner posted each year."
cc <- proj |>
  transmute(year, `Drop-out rate` = dnf_rate, `Par time at index 758` = par_h) |>
  pivot_longer(-year, names_to = "metric", values_to = "value") |>
  mutate(metric = factor(metric, levels = c("Drop-out rate", "Par time at index 758")),
         lab = if_else(metric == "Drop-out rate", percent(value, 1), fmt_hms(value * 3600)))

ggplot(cc, aes(year, value, fill = year)) +
  geom_col(width = 0.62) +
  geom_text(aes(label = lab), vjust = -0.4,
            family = "Helvetica Neue", colour = ft_ink, size = 3.4) +
  facet_wrap(~ metric, scales = "free_y") +
  scale_fill_manual(values = c(`2024` = ft_claret, `2025` = ft_orange, `2026` = ft_blue),
                    guide = "none") +
  scale_y_continuous(expand = expansion(c(0, 0.15))) +
  labs(title = "A deeper field or a harder day costs places",
       subtitle = "Attrition and the par time for a 758-index finisher, by edition.",
       x = NULL, y = NULL, caption = srcline())
```

::: {.callout-note appearance="simple"}
**Does a changed course break C and D?** For a *uniform* change — one that slows everyone by roughly
the same amount — no, and this is testable rather than assumed. If the year effect were uniform, the
index→time slope would be the same in every edition and only the intercept would move. It essentially
does: `r round(coef(y24$fit)[2] * 60, 2)`, `r round(coef(y25$fit)[2] * 60, 2)` and
`r round(coef(y26$fit)[2] * 60, 2)` min per index point over the fitted range, differences well
inside the noise. A formal test for an index × edition interaction gives
**p = `r sprintf("%.2f", int_p_restricted)`** over the fitted range and
**p = `r sprintf("%.2f", int_p_full)`** over the whole field. The year effect really is a clean level shift, which is exactly the shape C
and D assume — a point in their favour, and one worth testing rather than asserting.

**What they do not absorb.** Three things. A *non-uniform* course change would reshuffle who is fast
and break both. A change in the **index scale** between editions shifts the x-axis rather than the
line, and lens D does not see it — that is dealt with [below](#drift). And **who finishes** differs
by year (DNF rate `r percent(y24$dnf_rate,1)` / `r percent(y25$dnf_rate,1)` /
`r percent(y26$dnf_rate,1)`), which truncates the slow tail differently and moves the very intercept
D reads; restricting the fit to index `r FIT_LO`–`r FIT_HI` limits but does not eliminate this.
:::

# Is a 758 in 2026 a 758 in 2024? {#drift}

The whole exercise drops a **2026-vintage** index into fields rated on the **2024 and 2025** scales.
That is only legitimate if the scale has not moved.

The scrape does return each edition's contemporaneous index rather than today's value — among
runners matched across 2024 and 2026, only **`r round(pct_identical[["2024"]], 1)`%** carry an
identical number. So the scale can drift, and it does. Measured within runner on those who ran two
editions, the median index rose **+`r drift[["2024"]]` points from 2024 to 2026**
(n = `r n_ret[["2024"]]`) and **+`r drift[["2025"]]` points from 2025 to 2026**
(n = `r n_ret[["2025"]]`). At `r round(abs(coef(y26$fit)[2]) * 60, 1)` min per point, ten points is
about half an hour of finish time — not a rounding error.

Two things keep this from overturning the answer.

**It is a known quantity, and correcting for it moves him a little.** Expressing his 758 on each
year's own scale — `r TI - drift[["2024"]]` for 2024, `r TI - drift[["2025"]]` for 2025 — gives
seeds of `r ord(g("2024","seed_fin_adj"))` and `r ord(g("2025","seed_fin_adj"))` among finishers
instead of `r ord(g("2024","seed_fin"))` and `r ord(g("2025","seed_fin"))`, and moves lens C to
`r ord(g("2024","adj_drift"))` (2024) and `r ord(g("2025","adj_drift"))` (2025) — some
`r g("2024","adj_drift") - g("2024","adj_delta")` and
`r g("2025","adj_drift") - g("2025","adj_delta")` places worse. Both remain comfortably ahead of
`r ord(tangi_rank26)`, and both sit inside the region set out under
[How firm is this?](#firm). Note also that
returnees are a selected sample — runners who come back and get re-rated tend to be improving — so
this correction is an **upper bound** on pure scale inflation, not a best estimate.

**His own ratings say most of the rise was real.** Tangi's UTMB Index on race day 2025 was
**`r TI_2025_actual`**, against `r TI` in 2026: **+`r TI - TI_2025_actual` points** in a year, where
the population drifted **+`r drift[["2025"]]`**. At most half of his own improvement can be scale;
the rest is a genuinely stronger runner. The counterfactual this report projects — a 2026-calibre
Tangi in the 2025 field — is therefore a materially better runner than the 2025-vintage one, whether
the comparison is made at `r TI` or at the drift-adjusted `r TI - drift[["2025"]]`.

```{r drift-table}
tibble(
  `His index treated as` = c(
    paste0(TI_2025_actual, " (his actual 2025 rating)"),
    paste0(TI - drift[["2025"]], " / ", TI - drift[["2024"]], " (drift-adjusted)"),
    paste0(TI, " (2026 value, as used)")),
  `2024 seed` = c(ord(seed_fin(y24, TI_2025_actual)), ord(g("2024","seed_fin_adj")), ord(g("2024","seed_fin"))),
  `2025 seed` = c(ord(seed_fin(y25, TI_2025_actual)), ord(g("2025","seed_fin_adj")), ord(g("2025","seed_fin"))),
  `2026 seed` = c(ord(seed_fin(y26, TI_2025_actual)), ord(seed_fin(y26, TI)), ord(seedF26))
) |>
  kable(align = "lrrr",
        caption = "Sensitivity of the finisher-basis seed to the index scale. The drift-adjusted row uses each year's own median returnee drift.")
```

# Bottom line

```{r bottom-line}
#| output: asis
cat(paste0("- ", prior |>
  mutate(txt = paste0("**", year, "**: about ", ord(central),
                      " (", ord(spread_lo), "–", ord(spread_hi),
                      " across the three methods) — some ", tangi_rank26 - central,
                      " places ahead of his real ", ord(tangi_rank26),
                      ", in a field of ", comma(n_fin), " finishers.")) |>
  pull(txt), collapse = "\n"))
```

Run the same race in 2024 or 2025 and Tangi finishes around `r ord(g("2024","central"))` and
`r ord(g("2025","central"))` respectively — some `r tangi_rank26 - g("2024","central")` to
`r tangi_rank26 - g("2025","central")` places better than his real `r ord(tangi_rank26)`. The cause
is the field, not the runner: 2026 brought `r sum(y26$f$utmb_index > TI)` higher-rated finishers to
the line against `r sum(y24$f$utmb_index > TI)` and `r sum(y25$f$utmb_index > TI)` in those years. On
a like-for-like basis his 2026 run was an ordinary one for a `r TI` index, and that ordinary run is
what has been carried across.

Read as a region rather than a place, that is roughly
`r ord(firm("2024")[["lo"]])`–`r ord(firm("2024")[["hi"]])` in 2024 and
`r ord(firm("2025")[["lo"]])`–`r ord(firm("2025")[["hi"]])` in 2025. Two things stop it from being
sharper. His personal margin over par is
`r round(resid26 / resid_sd26, 2)` standard deviations, measured on **one race** — remove it
entirely and he still finishes `r ord(par_rank[["2024"]])` and `r ord(par_rank[["2025"]])`, so almost
all of the gain is the field and almost none of it is him. And the index scale itself drifted by
roughly `r drift[["2025"]]` points a year, which costs him about ten places once corrected. What
survives both is the direction and its rough size.

# Methodology {#methodology}

Both working lenses transplant Tangi's **2026 run** onto the 2024 and 2025 fields. His 2026 UTMB
Index is $I_T = `r TI`$; his real 2026 finish is $r_{26} = `r tangi_rank26`$ of `r comma(y26$n_fin)`
finishers. Each edition uses the full field (finishers + DNFs) restricted to runners carrying an
index.

## Bases

Everything is counted **among finishers**:

$$\text{seed}_y = 1 + \#\{\, i \in \text{finishers}_y : I_i > I_T \,\}.$$

Mixing a starter-basis seed with a finisher-basis rank is the error diagnosed in
[the form section](#form): it removes `r comma(y26$n_dnf)` runners from the denominator but not the
numerator and awards every finisher a spurious positive margin (median `r med_dS` places).

## C · Seed-adjusted (rank-based)

Form, measured like for like in 2026:

$$\rho_F = \frac{r_{26}}{\text{seed}_{26}} = \frac{`r tangi_rank26`}{`r seedF26`} = `r round(formF_ratio, 2)`,
\qquad
\delta_F = \text{seed}_{26} - r_{26} = `r formF_delta`\ \text{places},$$

projected as $\hat r^{\,\rho}_y = \rho_F \times \text{seed}_y$ and
$\hat r^{\,\delta}_y = \text{seed}_y - \delta_F$. C uses only ranks and ratings, never a clock, so
any change that slows the whole field equally leaves it untouched by construction. Its assumptions:
the index is comparable across editions ([tested](#drift)); form is a stable personal trait
(imposed by the question rather than tested — see [How firm is this?](#firm)); and the seed→finish
relationship keeps a similar shape.

**Robustness — is the finisher-basis seed a fluke of who dropped?** Fitting a per-edition logistic
model of finishing on a natural spline in index and summing the fitted probabilities over the
higher-rated starters gives the *expected* number of higher-rated finishers:

```{r dnf-model}
dnfm <- imap(ed, \(Y, yr) {
  m  <- glm(fin ~ ns(utmb_index, 4), data = Y$s, family = binomial)
  hi <- Y$s |> filter(utmb_index > TI)
  tibble(Edition = yr,
         `Higher-rated starters` = nrow(hi),
         `Expected to finish` = round(sum(predict(m, hi, type = "response")), 1),
         `Actually finished` = sum(hi$fin))
}) |> list_rbind()
kable(dnfm, align = "lrrr")
```

The realised counts sit within a few runners of the modelled expectation in every edition, so the
between-year differences in seed reflect the composition of the field rather than a lucky or unlucky
pattern of drop-outs.

## D · Index-residual time model

Fit $\widehat{\text{time}}_y(I) = a_y + b_y I$ on year $y$'s finishers, take his personal margin
$\varepsilon = \text{time}_T - \widehat{\text{time}}_{26}(I_T)$, which is `r mins(resid26)` faster
than par, and rank $\hat t_y = \widehat{\text{time}}_y(I_T) + \varepsilon$ against that year's finishers.

**Range restriction.** The fit uses index `r FIT_LO`–`r FIT_HI` rather than the whole field. Below
that range the sample is heavily truncated by the cut-off and the truncation differs by year (DNF
rate `r percent(y24$dnf_rate,1)` / `r percent(y25$dnf_rate,1)` / `r percent(y26$dnf_rate,1)`), which
flattens the bottom of the line and biases exactly the intercept D reads. The effect is material:
over the full range the par time at `r TI` is `r fmt_hms(g("2024","par_h_full")*3600)` (2024) against
`r fmt_hms(pred26_full*3600)` (2026), a gap of `r mins(g("2024","par_h_full") - pred26_full)`;
restricted, it is `r fmt_hms(g("2024","par_h")*3600)` against `r fmt_hms(pred26*3600)`, a gap of
`r mins(g("2024","par_h") - pred26)`. Roughly a quarter of the apparent "2024 was a slower year"
signal is selection. The full-range projections are shown alongside the restricted ones in the
[lens table](#the-lenses); they differ by `r abs(g("2024","model_full") - g("2024","model"))` places
in 2024 and `r abs(g("2025","model_full") - g("2025","model"))` in 2025.

**Assumption.** $\varepsilon$ is constant in absolute hours. If a year were much harder, a strong
runner's absolute advantage might grow, which D would under-count — so D is, if anything, slightly
conservative in the hardest year. Fit quality: $R^2 \approx `r round(summary(y26$fit)$r.squared, 2)`$ over the restricted range,
against `r round(summary(y26$fit_full)$r.squared, 2)` over the whole field — the drop is range
attenuation, not a worse model, and Tangi sits in the middle of the fitted range.

## The reported range, and what it is not

The range in the headline table is the **spread across the three methods** (C-delta, C-ratio, D). It
is reported because it shows the arithmetic is not knife-edged, and it is explicitly **not** a
confidence interval: three deterministic recipes sharing a single input cannot disagree about that
input, so their agreement measures nothing about how well the input is known.

The uncertainty that *can* be quantified from this data is sampling error in the two fitted lines
behind lens D. Bootstrapping both fits over the restricted range (600 replicates, resampling
finishers with replacement) gives `r ord(bd("2024","lo"))`–`r ord(bd("2024","hi"))` for 2024 and
`r ord(bd("2025","lo"))`–`r ord(bd("2025","hi"))` for 2025 — several times the spread across methods,
and still only one term. The others are set out in [How firm is this?](#firm): the thinness of the
transplanted margin (a zero-margin run costs him
`r par_rank[["2024"]] - g("2024","model")`–`r par_rank[["2025"]] - g("2025","model")` places), the
density of finishers around his projected time, and the chance of not finishing at all. None is
folded into a single published interval, because doing so would require assuming a distribution for
the one quantity — his form on a day he did not run — that is genuinely unknown.

## What is excluded

Raw time (B) is reported but not used: it is blind to conditions and to any course change, and it
implies he would have finished `r ord(g("2024","raw_time"))` in 2024 — better than every
field-strength-aware lens — purely because 2024 was a slower year. Seed (A) is reported as the input
to C. A *non-uniform*
course change — a materially more technical or higher route — would break both working lenses;
across 2024–2026 UTMB ran essentially the same classic loop
(`r round(173.6, 1)` km reconstructed in 2026 against 176.7 km in 2024 and 175.4 km in 2025), so this
is treated as second-order.

```{r write-csv}
#| include: false
proj |>
  left_join(boot_D |> select(year, boot_lo = lo, boot_hi = hi), by = "year") |>
  select(year, n_start, n_fin, dnf_rate, med_index,
         seed_start, seed_fin, seed_fin_adj,
         raw_time, adj_delta, adj_ratio, adj_drift, model, model_full,
         central, spread_lo, spread_hi, boot_lo, boot_hi,
         par_h, par_h_full, proj_time_h) |>
  write_csv("../data/processed/tangi_cross_year.csv")
```
